Making Deep Q-learning Methods Robust to Time Discretization

### 2019-01-19 paper/icml_drau/advup.tex

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
f(s),\,A(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same $Q$ function. This new $A$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A$ and $V$. This translates to
\begin{equation}
	A(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
		    &= A(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
\begin{equation}
\label{eq:Aparam}
	A(s, a) \deq \bar{A}(s, a) - \bar{A}(s, \pi(s)).
\end{equation}
With this parameterization, one can hope to learn an approximation of $A^\pi$,
from which a proper ranking of actions can be derived for any $\deltat$.

\subsection{Time invariant exploration}
\label{subsec:explo}
To obtain a time invariant RL algorithm, a time invariant
exploration scheme is required. 
\cite{ddpg} already introduces a time discretization invariant exploration scheme for continuous actions, which consists in adding an
\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions. Formally, it is defined as
\begin{equation}
	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) = \pi(s^k_\deltat) + z^k_\deltat
\end{equation}
with $z^k_\deltat$ the discretization of a time continuous OU process, 
\begin{equation}
	dz_t = - z_t \kappa dt + \sigma \dbt.
	\label{eq:orn_uhl}
\end{equation}
where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and $\sigma$ a noise scaling parameter. Resulting discretized trajectories converge
to non trivial continuous time trajectories.

This exploration can be extended to schemes of the form
\begin{equation}
  \label{eq:explore}
	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
\end{equation}
with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
A sufficient condition for the exploratory policy to admit a time continuous
limit is for
$z_\deltat$ to admit a limit, i.e. that the sequence $z_\deltat$ converges to a
well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
% For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
% being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
% and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
% \begin{equation}
% 	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
% \end{equation}}
To obtain a temporally coherent exploration scheme in discrete action environment
$z_\deltat$ is chosen to be a discretization of a $|\mathcal{A}|$ dimensional
continuous OU process, and set
\begin{equation}
  \pi^\text{explore}(s_t, z_t)= \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
\end{equation}
where $z_t[a']$ denotes the $a'$-th component of the discretized OU process at time
$k$. The resulting exploration scheme converges to a proper time continuous exploration scheme,
and is non trivial for most $Q$'s.

>>>>

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
f(s),\,A(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same $Q$ function. This new $A$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A$ and $V$. This translates to
\begin{equation}
	A(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed \TODO{keep or  leave as
an exercise?}
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
		    &= A(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
\begin{equation}
\label{eq:Aparam}
	A(s, a) \deq \bar{A}(s, a) - \bar{A}(s, \pi(s)).
\end{equation}
With this parameterization, one can hope to learn an approximation of $A^\pi$,
from which a proper ranking of actions can be derived for any $\deltat$.

\subsection{Timestep-invariant exploration}
\label{subsec:explo}

To obtain a timestep-invariant RL algorithm, a timestep-invariant
exploration scheme is required. 
For continuous actions, \cite{ddpg} already introduced a time discretization invariant exploration scheme, which consists in adding an
\emph{Ornstein--Uhlenbeck} \cite{orn-uhl} (OU) process to actions. Formally, it is defined as
\begin{equation}
	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) \deq \pi(s^k_\deltat) + z^k_\deltat
\end{equation}
with $z^k_\deltat$ the discretization of a continuous-time OU process,
\begin{equation}
	dz_t = - z_t \,\kappa\, dt + \sigma \,\dbt.
	\label{eq:orn_uhl}
\end{equation}
where $B_t$ is a brownian motion, $\kappa$ a stiffness parameter and
$\sigma$ a noise scaling parameter. The discretized trajectories converge
to non trivial continuous time trajectories, exhibiting a Brownian
behavior with a recall force towards $0$.

This exploration can be extended to schemes of the form
\begin{equation}
  \label{eq:explore}
	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
\end{equation}
with $(z^k_\deltat)_{k\geq 0}$ a sequence of random variables independent from the $a$'s and $s$'s.
A sufficient condition for this policy to admit a continuous-time
limit is for the sequence
$z_\deltat$ to converge in law to a
well-defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
% For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
% being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
% and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
% \begin{equation}
% 	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
% \end{equation}}

Thus, for discrete actions we can obtain a temporally consistent
exploration scheme by taking
$z_\deltat$ to be a discretization of an $|\mathcal{A}|$-dimensional
continuous OU process, and setting
\begin{equation}
  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(Q(s, a') + z_t[a']\right)
\end{equation}
where $z_t[a']$ denotes the $a'$-th component of $z_t$. Namely, we
perturb $Q$-values by a random process. The resulting exploration scheme
converges in continuous time,
and is non trivial for most $Q$'s.
